---
title: "Effect of car transmission type on MPG"
output: pdf_document
---

# Executive Summary

In this report, we are to (1) examine whether an automatic or manual 
transmission is better for MPG and (2) quantify the MPG difference between the
two types of transmissions. After performing a multiple linear regression to 
find the best-fit model, we conclude that the type of transmission does not
actually directly impact MPG.

# Data Loading & Preparation

The data is retrieved from the `mtcars` data set provided by the `datasets`
package.

```{r}
data(mtcars)
```

# Exploratory Data Analysis

The `mpg` data was plotted against all other variables (see Figure 1 in the 
appendix). An initial overview of the `mpg ~ am` boxplot seem to suggest that cars with
automatic transmissions seem to achieve less miles per gallon than cars with
manual transmissions. However, the plots of the other data also show some
correlation of `mpg` with the other variables available, suggesting the presence
of one or more confounding variables.

# Model fitting

## Variable selection

A measurement of each variable's correlation with `mpg` reveals that there is an
especially strong negative correlation with `cyl`, `disp` and `wt`:

```{r}
cor(mtcars$mpg, mtcars)
```

Based of this initial exploration, we will focus our efforts on `cyl`, `disp`
and `wt`.

## Model testing

```{r}
fit1 <- lm(mpg ~ am, data = mtcars)
fit2 <- lm(mpg ~ am + wt, data = mtcars)
fit3 <- lm(mpg ~ am + wt + cyl, data = mtcars)
fit4 <- lm(mpg ~ am + wt + cyl + disp, data = mtcars)
anova(fit1, fit2, fit3, fit4)
```

From this, we can see that `wt` and `cyl` are both significant whereas `disp` is
not, based on the p-values in a nested model Anova analysis with an alpha of
0.05.

Now, let's test the significance of `am` to our model:

```{r}
fit1 <- lm(mpg ~ wt + cyl, data = mtcars)
fit2 <- lm(mpg ~ wt + cyl + am, data = mtcars)
anova(fit1, fit2)
```

Interestingly, it appears that `am` is not significant, with a p-value 
significantly exceeding our alpha of 0.05. If we plot `mpg` against `wt` and 
highlight `cyl`  and `am`  in the data points, we see that both `cyl` and `am`
are proxy indicators of `wt`. The heavier a car, the more cylinders in the
engine (see Figure 2). Also, cars with automatic transmissions are heavier than
cars with manual transmissions (see Figure 3).

Therefore, the model most illustrative of effect on a car's MPG would be:

```{r}
summary(lm(mpg ~ wt, data = mtcars))$coef
```

See Figure 4 for residual and diagnostic plots.

# Conclusion

Based on our findings, we can conclude that manual transmissions are better than
automatic transmissions for MPG. However, this is not because of the transmission
type, but instead because of the weight of the car.

As for quantifying the MPG difference for weight, the true regressor, each 1000 
lb increase in weight of a car results in about 5.3 less MPG.

# Appendix

## Figure 1

```{r}
par(mfrow = c(2,3))
plot(mpg ~ ., data = mtcars)
```

## Figure 2

```{r}
plot(mtcars$wt, mtcars$mpg, pch=as.numeric(mtcars$cyl))
legend("topright", legend = levels(as.factor(mtcars$cyl)), pch=c(4,6,8))
```

## Figure 3

```{r}
plot(mtcars$wt, mtcars$mpg, pch = mtcars$am)
legend("topright", legend = c("automatic", "manual"), pch=as.numeric(levels(as.factor(mtcars$am))))
```

## Figure 4

```{r}
model <- lm(mpg ~ wt, data = mtcars)
par(mfrow = c(2, 2)); plot(model)
```
